---
title: "IPA illustration of Northern Tosk Albanian"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = here::here())
library(fs)
meta <- yaml::read_yaml("./meta.yaml")
```

`r badger::badge_custom("version", meta$repo_version, "green")`
`r badger::badge_custom("version", meta$repo_date, "blue")`
`r badger::badge_doi("10.17605/OSF.IO/VRY3H", "yellow")`

This is the GitHub repository of the project *IPA illustration of Northern Tosk Albanian*.
See the full **research compendium** at <https://osf.io/vry3h/>.

## License

Data is released under CC-BY 4.0 International, code is released under MIT License. See `LICENSE` file.

## Directory structure

- `docs/`: contains project documentation.
  - The file `procedure.md` documents data acquisition and processing in detail.
- `code/`: contains code for data (pre-)processing, organised in the following sub-directories.
  - `config/`: configuration script for setting up the EMU-SDMS database and the OFS repository.
  - `praat/`: Praat scripts for data-preprocessing.
  - `r/`: R scripts for data transformation, wrangling, and plotting.
- `img/`: figures generated by `code/prepare_plots.Rmd`.
- `renv/`: R package library managed by the renv R package.

## Data

The data of this project is version-controlled with [dvc](https://dvc.org).
A separate `dvc` repository---which is independent from this `git` repository---holds the data, and the `data.dvc` file, which is instead versioned by `git`, enables data transfer to and from the `dvc` repository.
Due to server limitations at the hosting institution, we cannot grant access to the `dvc` repository, and we instead make the data available as a `.zip` archive on the Open Science Framework (<https://osf.io/u43sg/>).

The archive contains the directories of this GitHub repository and the `data` directory with the project data.
The `data` directory contains the following sub-directories:

- `metadata/`: metadata files.
- `stimuli/`: contains the list of words (`word-list.txt`) and randomised versions of this list (`rand-1:6.txt`).
- `recordings/`: raw and derived recordings, used for forced-alignment and acoustic analysis.
  - `raw/`: raw (original) `.wav` recordings with accompanying `.TextGrids` used for chunking the recordings into separate parts.
  - `derived/`
    - `align/`: files generated by the forced-alignment procedure.
    - `post-align/`: post-alignment files, imported into the EMU database `alb-ipa_emuDB/`. To avoid data duplication, this folder is not version-controlled, so it doesn't ship with this repository.
    - `illustr/`: illustrative recordings of `consonants/`, `vowels/`, `intonation/`, and *The North Wind and the Sun*.
    - `examples/`: recordings for the figures with annotated waveform and spectrogram exemplifying VOT and frication.
- `alb-ipa_emuDB/`: EMU-SDMS database.
- `ultrasound/`: tongue contour spline data from ultrasound tongue imaging (UTI data on the OSF).
- `varia/`: a miscellanea folder with files used with different services.

## Data collection and processing

Data collection and processing are documented in `docs/procedure.md`.

## Directory tree

```{r tree}
dir_tree(recurse = 2, type = "directory")
```

