---
title: "IPA illustration of Northern Tosk"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = here::here())
library(fs)
```

This is the GitHub repository of the project *IPA illustration of Northern Tosk*.

## Directory structure

- `docs/`: contains project documentation.
  - The file `procedure.md` documents data acquisition and processing in detail.
- `code/`: contains code for data (pre-)processing, organised in the following sub-directories.
  - `config/`: configuration script for setting up the EMU-SDMS database.
  - `praat/`: Praat scripts for data-preprocessing.
  - `r/`: R scripts for data transformation, wrangling, and plotting.
- `img/`: figures generated by `code/prepare_plots.Rmd`.
- `renv/`: R package library managed by the renv R package.

## Data

The data of this project is version-controlled with [dvc](https://dvc.org).
A separate `dvc` repository---which is independent from this `git` repository---holds the data, and the `data.dvc` file, which is instead versioned by `git`, enables data transfer to and from the `dvc` repository.
Due to server limitations at the hosting institution, we cannot grant access to the `dvc` repository, and we instead make the data available as an archive (which also includes the contents of this GitHub repository) on the Open Science Framework (TBA).

The archive contains the directories of this GitHub repository and the `data` directory with the project data.
The `data` directory contains the following sub-directories:

- `metadata/`: metadata files.
- `stimuli/`: contains the list of words (`word-list.txt`) and randomised versions of this list (`rand-1:6.txt`).
- `recordings/`: raw and derived recordings, used for forced-alignment and acoustic analysis.
  - `raw/`: raw (original) `.wav` recordings with accompanying `.TextGrids` used for chunking the recordings into separate parts.
  - `derived/`
    - `align/`: files generated by the forced-alignment procedure.
    - `post-align/`: post-alignment files, imported into the EMU database `alb-ipa_emuDB/`. To avoid data duplication, this folder is not version-controlled, so it doesn't ship with this repository.
    - `illustr/`: illustrative recordings of `consonants/`, `vowels/`, `intonation/`, and *The North Wind and the Sun*.
    - `examples/`: recordings for the figures with annotated waveform and spectrogram exemplifying VOT and frication.
- `alb-ipa_emuDB/`: EMU-SDMS database.
- `ultrasound/`: tongue contour spline data from ultrasound tongue imaging (UTI data on the OSF).
- `varia/`: a miscellanea folder with files used with different services.

## Data collection and processing

Data collection and processing are documented in `docs/procedure.md`.

## Directory tree

```{r tree}
dir_tree(recurse = 2, type = "directory")
```

