---
title: "IPA illustration of Northern Tosk"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = here::here())
library(fs)
```

This is the GitHub repository of the project *IPA illustration of Northern Tosk*.

## Directory structure

- `docs/`: contains project documentation.
  - The file `procedure.md` documents data acquisition and processing in detail.
- `code/`: contains code for data (pre-)processing, organised in the following sub-directories.
  - `config/`: configuration script for setting up the EMU-SDMS database.
  - `praat/`: Praat scripts for data-preprocessing.
  - `r/`: R scripts for data transformation, wrangling, and plotting.
- `data/`: contains different sources of data. This folder is tracked with `dvc`.
  - `metadata/`: metadata files.
  - `stimuli/`: contains the list of words (`word-list.txt`) and randomised versions of this list (`rand-1:6.txt`).
  - `recordings/`: raw and derived recordings, used for forced-alignment and acoustic analysis.
    - `raw/`: raw (original) `.wav` recordings with accompanying `.TextGrids` used for chunking the recordings into separate parts.
    - `derived/`
      - `align/`: files generated by the forced-alignment procedure.
      - `post-align/`: post-alignment files, imported into the EMU database `alb-ipa_emuDB/`. To avoid data duplication, this folder is not version-controlled, so it doesn't ship with this repository.
      - `illustr/`: illustrative recordings of `consonants/`, `vowels/`, `intonation/`, and *The North Wind and the Sun*.
  - `alb-ipa_emuDB/`: EMU-SDMS database.
  - `ultrasound/`: spline data from ultrasound tongue imaging.
  - `varia/`: a miscellanea folder with files used with different services.
- `img/`: figures generated by `code/prepare_plots.Rmd`.
- `renv/`: R package library managed by the renv R package.

## Data collection and processing

Data collection and processing are documented in `docs/procedure.md`.

## Directory tree

```{r tree}
dir_tree(recurse = 2, type = "directory")
```

