---
title: "Configuration procedure of the emuDB"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
knitr::opts_knit$set(root.dir = here::here())
library(emuR)
library(wrassp)
library(speakr)
library(tools)

# utility function to run Praat
source("https://raw.githubusercontent.com/IPS-LMU/The-EMU-SDMS-Manual/master/R/praat2AsspDataObj.R")
```

# Create emuDB

To create an emuDB, we import the WebMAUS output files from `./data/recordings/derived/post-align/` into `./data/alb-ipa_emuDB/`.

The WebMAUS output TextGrids contain the following tiers:

- `ORT-MAU`: Written transcription.
- `KAN-MAU`: MAUS word canonical phonetic transcription.
- `MAU`: MAUS segment transcription.

These will be converted to `SEGMENT` (time-based) levels with the same names in the emuDB.

```{r tg-to-emu}
convert_TextGridCollection(
  "./data/recordings/derived/post-align/",
  "alb-ipa",
  "./data"
)
```

# Load emuDB

We load the newly created emuDB with `load_emuDB()`.
An SQLite connection to the database is opened and linked in `db_handle`.

```{r alb-ipa}
path <- "./data/alb-ipa_emuDB"

cat(crayon::yellow("\n\u2747 Checking emuDB cache...\n"))
db_handle <- load_emuDB(path, verbose = TRUE)

cat(crayon::green("\n\u2705 emuDB connection established!\n"))
db_handle$connection
```

# emuDB configuration

## Create hierarchical annotation

The EMU-SDMS allows for the creation of a time-less hierarchical annotation based on link definitions between levels.
The terminal nodes of the hierarchical annotation can be time-based levels, from which the times of the higher levels can be derived.
Time-less levels in the EMU-SDMS are of type `ITEM`.

`autobuild_linkFromTimes()` links a pair of time-based levels based on the intra-level time alignment of segment boundaries or events.
With the following code, we link `ORT-MAU` to `KAN-MAU` and `KAN-MAU` to `MAU`.
`ORT-MAU` and `KAN-MAU` are converted from `SEGMENT` (time-based) to `ITEM` (time-less) levels.

```{r autobuild}
autobuild_linkFromTimes(
  db_handle,
  superlevelName = "ORT-MAU",
  sublevelName = "KAN-MAU",
  newLinkDefType = "ONE_TO_MANY",
  convertSuperlevel = TRUE,
)

autobuild_linkFromTimes(
  db_handle,
  superlevelName = "KAN-MAU",
  sublevelName = "MAU",
  newLinkDefType = "ONE_TO_MANY",
  convertSuperlevel = TRUE,
)
```

## Add levels and link definitions

We add more levels used for annotation and create link definitions.

The following levels are added:

- `RELS` of type `EVENT`: time-point level for the annotation of stop-closure releases.
- `VOI` of type `EVENT`: time-point level for the annotation of the onset of vocal fold vibration following the stop-closure release.
- `SYL` of type `ITEM`: time-less level for the annotation of syllabic structure.
- `MAU2` of type `SEGMENT`: copy of `MAU` for the annotation of syllabic structure.

```{r add-levels}
add_levelDefinition(db_handle, "RELS", "EVENT")
add_levelDefinition(db_handle, "VOI", "EVENT")
add_levelDefinition(db_handle, "SYL", "ITEM")
```

We can now create new link definitions:

```{r add-link-defs}
add_linkDefinition(db_handle, "ONE_TO_ONE", "MAU", "RELS")
add_linkDefinition(db_handle, "ONE_TO_ONE", "MAU", "VOI")
add_linkDefinition(db_handle, "ONE_TO_MANY", "KAN-MAU", "SYL")
add_linkDefinition(db_handle, "ONE_TO_MANY", "SYL", "MAU2")
```

## Add track data

```{r}
wav_paths <- list.files(path, pattern = ".*wav$", recursive = T, full.names = T)

for (fp in wav_paths) {
  praatObj <- praat2AsspDataObj(fp, object = "formant")
  newPath <- paste0(file_path_sans_ext(fp), ".praatFms")
  write.AsspDataObj(praatObj, file = newPath)
}
```

