---
title: "Formants and VOT"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
library(emuR)
library(wrassp)
library(tidyverse)
```

# Prepare emuDB

```{r emudb}
alb_ipa <- "./data/alb-ipa_emuDB"
alb_int <- "./data/alb-ipa-int_emuDB"
dataframes <- "./data/dataframes"

alb_ipa_db <- load_emuDB(alb_ipa, verbose = TRUE)
alb_ipa_sql <- alb_ipa_db$connection
alb_ipa_items <- tbl(alb_ipa_sql, "items")
alb_ipa_links <- tbl(alb_ipa_sql, "links") %>% select(-label)
alb_ipa_labels <- tbl(alb_ipa_sql, "labels")

alb_ipa_int <- load_emuDB(alb_int, verbose = TRUE)
```

# Read participants metadata

```{r participants}
participants <- read_csv("./data/metadata/participants.csv") %>%
  select(-notes)
```

# IPA list 

```{r ipa-list}
ipa_list <- list(
  vowels_repl = list("@" = "\u025C"),
  vowels = list(i = "i", y = "y", u = "u", e = "e", roe = "\u025C", o = "o", a = "a"),
  cons_repl = list(
    "g" = "\u261",
    "D" = "\uF0", "T" = "\u3B8", "S" = "\u283", "Z" = "\u292",
    "tS" = "t\u283", "dZ" = "d\u29", "c" = "t\u33B\u283\u33B", "J" = "d\u33B\u292\u33B"
  ),
  plos = list("p", "t", "k", "b", "d", "\u261"),
  fric = list("f", "\u3B8", "s", "\u283", "h", "v", "\uF0", "z", "\u292"),
  affr = list("ts", "t\u283", "t\u33B\u283\u33B", "dz", "d\u292", "d\u33B\u292\u33B")
)
ipa_list$cons <- c(ipa_list$plos, ipa_list$fric, ipa_list$affr)
```

# Prepare data

## Formants

First, we query the emuDB to get the vowels in the relevant example words.

```{r vow}
vow <- alb_ipa_db %>%
  # Get vowels from the vowel-words in the MAU level
  query(
  "[MAU == p -> #MAU == e| i | y | u | @ | o | a]",
  bundlePattern = "[0-9]*-(peri|por|puro|pyk3|pik3|p3r|para)"
) %>%
  # Requery each vowel segment to get the word from the ORT-MAU level
  bind_cols(
    requery_hier(alb_ipa_db, ., "ORT-MAU") %>% select(labels) %>% rename(word = labels)
  ) %>%
  left_join(y = participants)
```

Now we can get formants from the midpoint of the vowels.

```{r formants}
formants <- get_trackdata(
  alb_ipa_db,
  seglist = vow,
  # Get values from mid-point
  cut = 0.5,
  npoints = 1,
  ssffTrackName = "FORMANTS",
  resultType = "tibble"
) %>%
  rename(
    F1 = T1,
    F2 = T2,
    F3 = T3,
    speaker = session
  ) %>%
  select(speaker, gender, labels, word, bundle, F1, F2, F3, start, end, sample_start, sample_end, times_orig) %>%
  mutate(
    vowel = str_replace_all(labels, unlist(ipa_list$vowels_repl)),
    vowel = factor(vowel, levels = ipa_list$vowels)
  ) %>%
  group_by(speaker) %>%
  mutate(
    F1.z = scale(F1),
    F2.z = scale(F2)
  ) %>%
  ungroup() %>%
  mutate(
    F1.z.hz = (F1.z * sd(F1)) + mean(F1),
    F2.z.hz = (F2.z * sd(F2)) + mean(F2)
  )
```

We now save the tibble.

```{r save-formants}
saveRDS(formants, here::here(dataframes, "formants.rds"))
```

## VOT

Query the database for the releases.

```{r vot}
vot <- alb_ipa_items %>%
  filter(level %in% c("RELS", "VOI")) %>%
  mutate(times = sample_point/sample_rate) %>%
  left_join(y = alb_ipa_links %>% rename(item_id = to_id)) %>%
  left_join(y = alb_ipa_labels %>% rename(from_id = item_id)) %>%
  select(session, bundle, level, times, label) %>%
  pivot_wider(names_from = "level", values_from = "times") %>%
  as_tibble() %>%
  mutate(
    vot = (VOI - RELS) * 1000,
    consonant = str_replace_all(label, unlist(ipa_list$cons_repl)),
    consonant = factor(consonant, levels = c(ipa_list$plos, ipa_list$affr))
  ) %>%
  rename(speaker = session)
```


```{r save-vot}
saveRDS(vot, here::here(dataframes, "vot.rds"))
```

## F0

First, we query the emuDB to get the relevant sentences.

```{r vow-f0}
vow_f0 <- alb_ipa_db %>%
  # Get vowels from the intonation-sentences in the MAU level
  query(
    "MAU !~ <p:>",
    bundlePattern = "[0-9]*-(broad*|polar*|narrow*)"
  ) %>%
  # Requery each vowel segment to get the word from the ORT-MAU level
  bind_cols(
    requery_hier(alb_ipa_db, ., "ORT-MAU") %>% select(labels) %>% rename(word = labels)
  )
```

```{r f0}
f0 <- alb_ipa_db %>%
  get_trackdata(
    vow_f0,
    ssffTrackName = "F0"
  ) %>%
  unite("token", session, bundle, remove = FALSE) %>%
  group_by(token) %>%
  mutate(
    sentence_dur = max(end) - min(start),
    norm_timepoint = ((times_orig - (min(start)))/sentence_dur),
    sentence_type = factor(substring(bundle, 5)),
    sentence = recode(
      sentence_type,
      "broad-focus" = "(a) Lena lau murin.",
      "narrow-focus" = "(b) Lena lau murin, jo lulen.",
      "polar-q" = "(c) A e lau Lena murin?",
      "narrow-focus-q" = "(d) Çfarë lau Lena?"
    )
  ) %>%
  left_join(y = participants) %>%
  rename(
    "f0" = "T1",
    "segment" = "labels",
    "speaker" = "session"
  )

f0 <- f0 %>%
  left_join(
    y = f0 %>%
      select(speaker, sentence, token, sentence_type) %>%
      unique() %>%
      ungroup() %>%
      mutate(number = 1) %>%
      group_by(speaker, sentence_type) %>%
      mutate(repetition = cumsum(number)) %>%
      ungroup() %>%
      select(speaker, sentence, token, repetition)
  ) %>%
  filter(
    f0 != 0,
    !(segment %in% c("tS", "f"))
  ) %>%
  select(speaker, gender, segment, word, sentence, sentence_type, repetition, f0, norm_timepoint, times_orig, sentence_dur, token)
```


And we now save the tibble.

```{r save-f0}
saveRDS(f0, here::here(dataframes, "f0.rds"))
```

## Intonation

```{r syl-f0}
syl <- query(alb_int_db, "SYL =~ .*", calcTimes = T) %>%
  filter(labels != "<p:>")

syl_f0 <- get_trackdata(alb_int_db, seglist = syl, ssffTrackName = "F0", cut = 0.5) %>%
  group_by(session, bundle) %>%
  mutate(
    sentence_dur = max(end) - min(start),
    norm_timepoint = ((times_orig - (min(start)))/sentence_dur)
  ) %>%
  rename(f0 = "T1", speaker = "session") %>%
  mutate(
    sentence_type = str_sub(bundle, 5),
    sentence = recode(
      sentence_type,
      "broad-focus" = "(a) Lena lau murin.",
      "narrow-focus" = "(b) Lena lau murin, jo lulen.",
      "polar-q" = "(c) A e lau Lena murin?",
      "narrow-focus-q" = "(d) Çfarë lau Lena?"
      )
  )
syl_f0[syl_f0$labels == "çfar(ë)",]$f0 <- 250

syl <- syl %>%
  group_by(session, bundle) %>%
  mutate(
    sentence_duration = max(end) - min(start),
    start_norm = (start - min(start)) / sentence_duration,
    end_norm = (end - min(start)) / sentence_duration,
    sentence_type = str_sub(bundle, 5),
    sentence = recode(
      sentence_type,
      "broad-focus" = "(a) Lena lau murin.",
      "narrow-focus" = "(b) Lena lau murin, jo lulen.",
      "polar-q" = "(c) A e lau Lena murin?",
      "narrow-focus-q" = "(d) Çfarë lau Lena?"
      )
  )
```

```{r save-syl-f0}
saveRDS(syl_f0, here::here(dataframes, "syl_f0.rds"))
saveRDS(syl, here::here(dataframes, "syl.rds"))
```


## Spectral moments (CoG)

### Fricatives

```{r fric}
# Get query list
fric_segs <- query(
  alb_ipa_db,
  "[MAU == f|v|T|D|s|z|S|Z|h]",
  bundlePattern = "[0-9]*-(fati|vata|that3|dhat3|salla|zalli|shava|zhaba|han3)"
)
```

```{r fric-dft}
# Get dft in wide format
fric_dft <- get_trackdata(
  alb_ipa_db,
  seglist = fric_segs,
  ssffTrackName = "dft",
  resultType = "tibble",
  cut = 0.5
)

fric_dft_l <- convert_wideToLong(fric_dft, calcFreqs = T)
```

```{r fric-moments}
# Get spectral moments
fric_moments <- fric_dft_l %>%
  filter(freq < 10000) %>%
  group_by(labels, sl_rowIdx) %>%
  do(data_frame(moments = moments(.$track_value,.$freq, minval = TRUE))) %>%
  mutate(moment_num = paste0("moment_", 1:(table(sl_rowIdx)))) %>%
  pivot_wider(names_from = "moment_num", values_from = "moments")

fric_moments %>%
  group_by(labels) %>%
  summarise(mean = mean(moment_1), sd = sd(moment_1))
```

```{r save-fric-dft-moments}
saveRDS(fric_dft_l, file = "./data/dataframes/fric_dtf_l.rds")
saveRDS(fric_moments, file = "./data/dataframes/fric_moments.rds")
```

### Affricates

```{r affr}
# Get query list
affr_rels = query(
  alb_ipa_db,
  "RELS =~ .*",
  bundlePattern = "[0-9]*-(cica|xixa|5aji|xhaja|qava|gjaku)", 
  calcTimes = T)

affr_segs = requery_hier(
  alb_ipa_db,
  affr_rels,
  level = 'MAU')

# Replace start of segments with time of release
affr_segs$start = affr_rels$start
```

```{r affr-dft}
# Get data in wide format
affr_dft = get_trackdata(
  alb_ipa_db,
  seglist = affr_segs,
  ssffTrackName = "dft",
  resultType = "tibble",
  cut = 0.5)


affr_dft_l = convert_wideToLong(affr_dft, calcFreqs = T)
```

```{r affr-moments}
# Get spectral moments
affr_moments = affr_dft_l %>%
  filter(freq < 10000) %>%
  group_by(labels, sl_rowIdx) %>%
  do(data_frame(moments = moments(.$track_value,.$freq, minval = TRUE))) %>%
  mutate(moment_num = paste0("moment_", 1:(table(sl_rowIdx)))) %>%
  pivot_wider(names_from = "moment_num", values_from = "moments")

affr_moments %>%
  group_by(labels) %>%
  summarise(mean = mean(moment_1), sd = sd(moment_1))
```


```{r save-affr-dft-moments}
# save dataframes
saveRDS(affr_dft_l, file = "./data/dataframes/affr_dft_l.rds")
saveRDS(affr_moments, file = "./data/dataframes/affr_moments.rds")
```


## Examples

```{r pata, eval=F}
pata <- query(
  alb_ipa_db,
  "[ORT-MAU == pata]",
  bundlePattern = "018-pata",
  sessionPattern = "s04"
)

export_seglistToTxtCollection(db, pata, "./data/varia/vot-example/")
```

```{r qgj, eval=FALSE}
qgj <- query(
  alb_ipa_db,
  "[ORT-MAU == gjaku|qava]",
  bundlePattern = "(044|038).*",
  sessionPattern = "s04"
)

export_seglistToTxtCollection(db, qgj, "./data/varia/qgj-example/")
```

